---
title: "Alignment in Growth Modeling: Neuroticism Example Using MIDUS"
date: "January 2, 2021, Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: true
  github_document: 
    html_preview: false
    toc: true
always_allow_html: true
header-includes:
  - \usepackage{booktabs}    
---

```{r midus-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(knitr.table.format = "latex")
```

## Load Required Packages

```{r load-packages, message=FALSE}
library(here) # for setting working directory
library(modelsummary) # for descriptive statistics
library(lavaan) # for fitting SEM model
library(sirt) # for alignment algorithm
```

## Download and Import Data

First, download the MIDUS data from the following links (registration on ICPSUR is needed):

- Wave I: https://www.icpsr.umich.edu/web/NACDA/studies/2760
- Wave II: https://www.icpsr.umich.edu/web/NACDA/studies/4652
- Wave III: https://www.icpsr.umich.edu/web/NACDA/studies/36346

For each zip file, extract the .rda data file from the "DS0001" folder. Specifically, the data file names are

- Wave I: "02760-0001-Data.rda"
- Wave II:  "04652-0001-Data.rda"
- Wave III: "36346-0001-Data.rda"

```{r import-midus}
# Import data (assuming rda data files are in the data directory)
midus1_name <- load(here::here("data", "02760-0001-Data.rda"))
midus2_name <- load(here::here("data", "04652-0001-Data.rda"))
midus3_name <- load(here::here("data", "36346-0001-Data.rda"))
# Select MIDUS 1 variables (ID, age, neuroticism items)
midus1_df <- get(midus1_name)[, c(
  "M2ID", "A1PRAGE_2019",
  "A1SF4C", "A1SF4H",
  "A1SF4M", "A1SF4S"
)]
# Select MIDUS 2 variables (ID, neuroticism items)
midus2_df <- get(midus2_name)[, c(
  "M2ID",
  "B1SE6C", "B1SE6H",
  "B1SE6M", "B1SE6S"
)]
# Select MIDUS 2 variables (ID, neuroticism items)
midus3_df <- get(midus3_name)[, c(
  "M2ID",
  "C1SE6C", "C1SE6H",
  "C1SE6M", "C1SE6S"
)]
# Merge data
midus12_df <- merge(midus1_df,
  y = midus2_df,
  by.x = "M2ID", by.y = "M2ID",
  all.x = TRUE, all.y = FALSE
)
midus123_df <- merge(midus12_df, midus3_df,
  by.x = "M2ID", by.y = "M2ID",
  all.x = TRUE, all.y = FALSE
)
```

### Subset Data

```{r midus_neurotic}
# Extract Neuroticism variables, and rename
midus_neurotic <-
  with(
    midus123_df,
    data.frame(
      m2id = M2ID,
      age = A1PRAGE_2019,
      moody1 = A1SF4C,
      worry1 = A1SF4H,
      nervous1 = A1SF4M,
      calm1 = A1SF4S,
      moody2 = B1SE6C,
      worry2 = B1SE6H,
      nervous2 = B1SE6M,
      calm2 = B1SE6S,
      moody3 = C1SE6C,
      worry3 = C1SE6H,
      nervous3 = C1SE6M,
      calm3 = C1SE6S
    )
  )
# Subset participants aged 55 or above in Wave 1
midus_neurotic <- subset(midus_neurotic, age <= 40)
# Drop missing data
midus_neurotic <- na.omit(midus_neurotic)
# Convert items from factor to numeric
midus_neurotic[, 3:14] <-
  lapply(midus_neurotic[, 3:14], as.numeric)
# Recode so that higher scores indicate higher neuroticism
# (4 to 1, 3 to 2, 2 to 3, 1 to 4)
midus_neurotic[c(
  "moody1", "worry1", "nervous1",
  "moody2", "worry2", "nervous2",
  "moody3", "worry3", "nervous3"
)] <-
  5 - midus_neurotic[c(
    "moody1", "worry1", "nervous1",
    "moody2", "worry2", "nervous2",
    "moody3", "worry3", "nervous3"
  )]
```

## Descriptive Statistics

```{r desc}
modelsummary::datasummary(
  (moody1 + worry1 + nervous1 + calm1 +
    moody2 + worry2 + nervous2 + calm2 +
    moody3 + worry3 + nervous3 + calm3) ~
  (Mean + SD + Median + Min + Max),
  data = midus_neurotic
)
```


## Longitudinal Configural Invariance Model

### `lavaan` script

```{r config_mod}
config_mod <- "
    eta1 =~ (lam11) * moody1 + (lam21) * worry1 + (lam31) * nervous1 +
            (lam41) * calm1
		eta2 =~ (lam12) * moody2 + (lam22) * worry2 + (lam32) * nervous2 +
		        (lam42) * calm2
		eta3 =~ (lam13) * moody3 + (lam23) * worry3 + (lam33) * nervous3 +
		        (lam43) * calm3
		# Measurement intercepts
    moody1 ~ (nu11) * 1
    worry1 ~ (nu21) * 1
    nervous1 ~ (nu31) * 1
    calm1 ~ (nu41) * 1
    moody2 ~ (nu12) * 1
    worry2 ~ (nu22) * 1
    nervous2 ~ (nu32) * 1
    calm2 ~ (nu42) * 1
    moody3 ~ (nu13) * 1
    worry3 ~ (nu23) * 1
    nervous3 ~ (nu33) * 1
    calm3 ~ (nu43) * 1
    # Unique factor covariances
    moody1 ~~ moody2 + moody3
    moody2 ~~ moody3
    worry1 ~~ worry2 + worry3
    worry2 ~~ worry3
    nervous1 ~~ nervous2 + nervous3
    nervous2 ~~ nervous3
    calm1 ~~ calm2 + calm3
    calm2 ~~ calm3
"
```

### Fit model in `lavaan`

```{r config_fit}
# Use lavaan::cfa() to fit a longitudinal configural model
config_fit <- cfa(
  config_mod,
  data = midus_neurotic,
  std.lv = TRUE # identify model by standardizing latent factors
)
# Uncomment to show model summary
# summary(config_fit, fit.measures = TRUE, standardized = TRUE)
```

## Longitudinal Alignment Optimization

```{r aligned_pars}
# Extract loadings and intercepts for alignment
lam_mat <- lavInspect(config_fit, what = "est")$lambda
nu_vec <- lavInspect(config_fit, what = "est")$nu
# Put them into T x p matrices
num_items <- 4
num_waves <- 3
lam_config <- crossprod(lam_mat, rep(1, num_waves) %x% diag(num_items))
nu_config <- matrix(nu_vec, nrow = num_waves, ncol = num_items, byrow = TRUE)
# Add indicator names
colnames(lam_config) <- colnames(nu_config) <-
  c("moody", "worry", "nervous", "calm")
# Alignment optimization
aligned_pars <- sirt::invariance.alignment(
  lambda = lam_config,
  nu = nu_config,
  fixed = TRUE
)
```

The aligned loadings and intercepts are shown below:

```{r tab-aligned_pars, echo=FALSE}
opts <- options(knitr.kable.NA = "")
library(magrittr)
library(kableExtra)
knitr::kable(
  rbind(
    "Loadings" = rep(NA, 4),
    `rownames<-`(aligned_pars$lambda.aligned,
                 paste0("Time", 1:3)),
    "Intercepts" = rep(NA, 4),
    `rownames<-`(aligned_pars$nu.aligned,
                 paste0("Time", 1:3))
  ),
  format = "simple",
  digits = 3L
)
```

<!-- ### $R^2$ Effect Size -->

```{r}
# lam_resid <- lam_config - tcrossprod(aligned_pars$pars[ , 2],
#                                      aligned_pars$itempars.aligned$M.lambda)
# 1 - apply(lam_resid, 2, var) / apply(lam_config, 2, var)
# Function for dMACS
dmacs <- function(loadings, intercepts, pooled_item_sd,
                  latent_mean = 0, latent_var = 1) {
  dloading <- diff(loadings)
  dintercept <- diff(intercepts)
  integral <- dintercept^2 + 2 * dintercept * dloading * latent_mean +
    dloading^2 * (latent_var + latent_mean^2)
  sqrt(integral) / pooled_item_sd
}
# Use item SDs at first time point
item_sds_wave1 <-
  apply(
    midus_neurotic[c("moody1", "worry1", "nervous1", "calm1")],
    2, sd
  )
dmacs_pairwise <- function(loading_mat, intercept_mat, pooled_item_sd,
                           latent_mean = 0, latent_var = 1) {
  ngroups <- nrow(loading_mat)
  pairs <- combn(ngroups, 2)
  out <- matrix(NA, nrow = ncol(pairs), ncol = ncol(loading_mat))
  for (i in seq_len(ncol(pairs))) {
    out[i, ] <- dmacs(loading_mat[pairs[, i], ],
      intercepts = intercept_mat[pairs[, i], ],
      pooled_item_sd,
      latent_mean,
      latent_var
    )
  }
  rownames(out) <- apply(pairs, 2, paste, collapse = " vs ")
  colnames(out) <- colnames(loading_mat)
  out
}
# All pairwise dMACS
dmacs_pairwise(aligned_pars$lambda.aligned,
  intercept_mat = aligned_pars$nu.aligned,
  pooled_item_sd = item_sds_wave1,
  latent_mean = 0,
  latent_var = 1
)
```

### An equivalent configural model with aligned loadings:

```{r config_mod2}
config_mod2 <- "
    # (First loadings fixed to alignment solution)
    eta1 =~ 0.4429136 * moody1 + (lam21) * worry1 + (lam31) * nervous1 +
            (lam41) * calm1
		eta2 =~ 0.4461272 * moody2 + (lam22) * worry2 + (lam32) * nervous2 +
		        (lam42) * calm2
		eta3 =~ 0.4615046 * moody3 + (lam23) * worry3 + (lam33) * nervous3 +
		        (lam43) * calm3
		# (First intercepts fixed to alignment solution)
    moody1 ~ 2.600240 * 1
    worry1 ~ (nu21) * 1
    nervous1 ~ (nu31) * 1
    calm1 ~ (nu41) * 1
    moody2 ~ 2.682681 * 1
    worry2 ~ (nu22) * 1
    nervous2 ~ (nu32) * 1
    calm2 ~ (nu42) * 1
    moody3 ~ 2.792330 * 1
    worry3 ~ (nu23) * 1
    nervous3 ~ (nu33) * 1
    calm3 ~ (nu43) * 1
    # Unique factor covariances
    moody1 ~~ moody2 + moody3
    moody2 ~~ moody3
    worry1 ~~ worry2 + worry3
    worry2 ~~ worry3
    nervous1 ~~ nervous2 + nervous3
    nervous2 ~~ nervous3
    calm1 ~~ calm2 + calm3
    calm2 ~~ calm3
    # Free latent means
    eta1 + eta2 + eta3 ~ NA*1
"
```

```{r config_fit2}
# Use lavaan::cfa() to fit a longitudinal configural model
config_fit2 <- cfa(
  config_mod2,
  data = midus_neurotic
)
# Uncomment to show model summary
# summary(config_fit2, fit.measures = TRUE, standardized = TRUE)
```

Compare model fit (they're equivalent)

```{r compare-config_fit_1vs2, warning = FALSE}
anova(config_fit, config_fit2)
```


## Alignment-Within-CFA (AwC) Approach for Growth Modeling

```{r awc_growth_mod}
awc_growth_mod <- "
        # (First loadings fixed to alignment solution)
    eta1 =~ 0.4429136 * moody1 + (lam21) * worry1 + (lam31) * nervous1 +
            (lam41) * calm1
		eta2 =~ 0.4461272 * moody2 + (lam22) * worry2 + (lam32) * nervous2 +
		        (lam42) * calm2
		eta3 =~ 0.4615046 * moody3 + (lam23) * worry3 + (lam33) * nervous3 +
		        (lam43) * calm3
		# (First intercepts fixed to alignment solution)
    moody1 ~ 2.399760 * 1
    worry1 ~ (nu21) * 1
    nervous1 ~ (nu31) * 1
    calm1 ~ (nu41) * 1
    moody2 ~ 2.317319 * 1
    worry2 ~ (nu22) * 1
    nervous2 ~ (nu32) * 1
    calm2 ~ (nu42) * 1
    moody3 ~ 2.207670 * 1
    worry3 ~ (nu23) * 1
    nervous3 ~ (nu33) * 1
    calm3 ~ (nu43) * 1
    # Unique factor covariances
    moody1 ~~ moody2 + moody3
    moody2 ~~ moody3
    worry1 ~~ worry2 + worry3
    worry2 ~~ worry3
    nervous1 ~~ nervous2 + nervous3
    nervous2 ~~ nervous3
    calm1 ~~ calm2 + calm3
    calm2 ~~ calm3
    # Linear Growth Model
    i =~ 1 * eta1 + 1 * eta2 + 1 * eta3
    s =~ 0 * eta1 + 1 * eta2 + 2 * eta3
    # Variance-covariances of intercepts and slopes
    i ~~ i
    s ~~ s
    i ~~ s
    # Means of level and slope
    i ~ 1
    s ~ 1
    # Fixed disturbances of latent outcomes to zero
    eta1 ~ 0 * 1
    eta2 ~ 0 * 1
    eta3 ~ 0 * 1
"
```

### Fit model in `lavaan`

```{r awc_growth_fit}
# Use lavaan::sem() to fit a second-order growth model
awc_growth_fit <- sem(
  awc_growth_mod,
  data = midus_neurotic
)
# Uncomment to show model summary
# summary(awc_growth_fit, fit.measures = TRUE, standardized = TRUE)
```

### Parameter Estimates

```{r pe-awc_growth_fit}
lavaan::parameterEstimates(awc_growth_fit) %>%
  subset(
    lhs %in% c("i", "s") & substr(rhs, 1, 3) != "eta",
    -label
  ) %>%
  knitr::kable(format = "simple", digits = 3L)
```

## Sensitivity Check: Identification Using Second Indicator

```{r awc_growth_mod2}
awc_growth_mod2 <- "
    # (First loadings fixed to alignment solution)
    eta1 =~ NA * moody1 + (lam11) * moody1 + 0.7855913 * worry1 +
            (lam31) * nervous1 + (lam41) * calm1
		eta2 =~ NA * moody2 + (lam12) * moody2 + 0.7975706 * worry2 +
		        (lam32) * nervous2 + (lam42) * calm2
		eta3 =~ NA * moody3 + (lam13) * moody3 + 0.7736944 * worry3 +
		        (lam33) * nervous3 + (lam43) * calm3
		# (First intercepts fixed to alignment solution)
    moody1 ~ (nu11) * 1
    worry1 ~ 2.618247 * 1
    nervous1 ~ (nu31) * 1
    calm1 ~ (nu41) * 1
    moody2 ~ (nu12) * 1
    worry2 ~ 2.629519 * 1
    nervous2 ~ (nu32) * 1
    calm2 ~ (nu42) * 1
    moody3 ~ (nu13) * 1
    worry3 ~ 2.617024 * 1
    nervous3 ~ (nu33) * 1
    calm3 ~ (nu43) * 1
    # Unique factor covariances
    moody1 ~~ moody2 + moody3
    moody2 ~~ moody3
    worry1 ~~ worry2 + worry3
    worry2 ~~ worry3
    nervous1 ~~ nervous2 + nervous3
    nervous2 ~~ nervous3
    calm1 ~~ calm2 + calm3
    calm2 ~~ calm3
    # Linear Growth Model
    i =~ 1 * eta1 + 1 * eta2 + 1 * eta3
    s =~ 0 * eta1 + 1 * eta2 + 2 * eta3
    # Variance-covariances of intercepts and slopes
    i ~~ i
    s ~~ s
    i ~~ s
    # Means of level and slope
    i ~ 1
    s ~ 1
    # Fixed disturbances of latent outcomes to zero
    eta1 ~ 0 * 1
    eta2 ~ 0 * 1
    eta3 ~ 0 * 1
"
```

### Fit model in `lavaan`

```{r awc_growth_fit2}
# Use lavaan::sem() to fit a second-order growth model
awc_growth_fit2 <- sem(
  awc_growth_mod2,
  data = midus_neurotic
)
# Uncomment to show model summary
# summary(awc_growth_fit, fit.measures = TRUE, standardized = TRUE)
```

### Parameter Estimates

```{r pe-awc_growth_fit2}
lavaan::parameterEstimates(awc_growth_fit2) %>%
  subset(
    lhs %in% c("i", "s") & substr(rhs, 1, 3) != "eta",
    -label
  ) %>%
  knitr::kable(format = "simple", digits = 3L)
```

Note the smaller standard errors when using item 2 as the reference indicator, which has larger loadings. 

## Traditional Invariance Testing

### Weak and Strong Invariance

```{r weak_mod}
# Weak invariance model
weak_mod <- "
    eta1 =~ NA * moody1 + (lam1) * moody1 + (lam2) * worry1 + (lam3) * nervous1 + 
            (lam4) * calm1
		eta2 =~ NA * moody2 + (lam1) * moody2 + (lam2) * worry2 + (lam3) * nervous2 + 
		        (lam4) * calm2
		eta3 =~ NA * moody3 + (lam1) * moody3 + (lam2) * worry3 + (lam3) * nervous3 + 
		        (lam4) * calm3
		# Measurement intercepts
    moody1 ~ (nu11) * 1
    worry1 ~ (nu21) * 1
    nervous1 ~ (nu31) * 1
    calm1 ~ (nu41) * 1
    moody2 ~ (nu12) * 1
    worry2 ~ (nu22) * 1
    nervous2 ~ (nu32) * 1
    calm2 ~ (nu42) * 1
    moody3 ~ (nu13) * 1
    worry3 ~ (nu23) * 1
    nervous3 ~ (nu33) * 1
    calm3 ~ (nu43) * 1
    # Unique factor covariances
    moody1 ~~ moody2 + moody3
    moody2 ~~ moody3
    worry1 ~~ worry2 + worry3
    worry2 ~~ worry3
    nervous1 ~~ nervous2 + nervous3
    nervous2 ~~ nervous3
    calm1 ~~ calm2 + calm3
    calm2 ~~ calm3
    # First factor variance to 1
    eta1 ~~ 1 * eta1
"
```

```{r weak_fit}
# Use lavaan::cfa() to fit a longitudinal configural model
weak_fit <- cfa(
  weak_mod, 
  data = midus_neurotic
)
# Uncomment to show model summary
# summary(weak_fit, fit.measures = TRUE, standardized = TRUE)
```

```{r strong_mod}
# Strong invariance model
strong_mod <- "
    eta1 =~ NA * moody1 + (lam1) * moody1 + (lam2) * worry1 + (lam3) * nervous1 + 
            (lam4) * calm1
		eta2 =~ NA * moody2 + (lam1) * moody2 + (lam2) * worry2 + (lam3) * nervous2 + 
		        (lam4) * calm2
		eta3 =~ NA * moody3 + (lam1) * moody3 + (lam2) * worry3 + (lam3) * nervous3 + 
		        (lam4) * calm3
		# Measurement intercepts
    moody1 ~ (nu1) * 1
    worry1 ~ (nu2) * 1
    nervous1 ~ (nu3) * 1
    calm1 ~ (nu4) * 1
    moody2 ~ (nu1) * 1
    worry2 ~ (nu2) * 1
    nervous2 ~ (nu3) * 1
    calm2 ~ (nu4) * 1
    moody3 ~ (nu1) * 1
    worry3 ~ (nu2) * 1
    nervous3 ~ (nu3) * 1
    calm3 ~ (nu4) * 1
    # Unique factor covariances
    moody1 ~~ moody2 + moody3
    moody2 ~~ moody3
    worry1 ~~ worry2 + worry3
    worry2 ~~ worry3
    nervous1 ~~ nervous2 + nervous3
    nervous2 ~~ nervous3
    calm1 ~~ calm2 + calm3
    calm2 ~~ calm3
    # First factor variance and mean to 1 and 0
    eta1 ~~ 1 * eta1
    eta1 ~ 0
    # Other factor means to free
    eta2 ~ NA * 1
    eta3 ~ NA * 1
"
```

```{r strong_fit}
# Use lavaan::cfa() to fit a longitudinal configural model
strong_fit <- cfa(
  strong_mod, 
  data = midus_neurotic
)
# Uncomment to show model summary
# summary(strong_fit, fit.measures = TRUE, standardized = TRUE)
```

```{r compare-fit}
fit_tab <- lapply(
  list("Configural" = config_fit, 
       "Weak" = weak_fit, 
       "Strong" = strong_fit),
  fitmeasures,
  fit.measures = c(
    "chisq", "df", "rmsea",
    "rmsea.ci.lower", "rmsea.ci.upper", 
    "cfi", "tli", "srmr"
  )
)
fit_tab <- do.call(rbind, fit_tab)
knitr::kable(fit_tab, format = "simple", digits = 3)
```

Using modification indices, two items were identified to have noninvariant intercepts: `moody` in Wave 3 and `calm` in Wave 1.

```{r mod}
modindices(strong_fit, sort. = TRUE, free.remove = FALSE, min = 10)
```

### Partial Strong Invariance Model

```{r pstrong_growth_fit}
pstrong_growth_mod <- "
    eta1 =~ NA * moody1 + (lam1) * moody1 + (lam2) * worry1 + (lam3) * nervous1 +
            (lam4) * calm1
		eta2 =~ NA * moody2 + (lam1) * moody2 + (lam2) * worry2 + (lam3) * nervous2 +
		        (lam4) * calm2
		eta3 =~ NA * moody3 + (lam1) * moody3 + (lam2) * worry3 + (lam3) * nervous3 +
		        (lam4) * calm3
		# Measurement intercepts
    moody1 ~ (nu1) * 1
    worry1 ~ (nu2) * 1
    nervous1 ~ (nu3) * 1
    calm1 ~ (nu41) * 1
    moody2 ~ (nu1) * 1
    worry2 ~ (nu2) * 1
    nervous2 ~ (nu3) * 1
    calm2 ~ (nu4) * 1
    moody3 ~ (nu13) * 1
    worry3 ~ (nu2) * 1
    nervous3 ~ (nu3) * 1
    calm3 ~ (nu4) * 1
    # Unique factor covariances
    moody1 ~~ moody2 + moody3
    moody2 ~~ moody3
    worry1 ~~ worry2 + worry3
    worry2 ~~ worry3
    nervous1 ~~ nervous2 + nervous3
    nervous2 ~~ nervous3
    calm1 ~~ calm2 + calm3
    calm2 ~~ calm3
    # Linear Growth Model
    i =~ 1 * eta1 + 1 * eta2 + 1 * eta3
    s =~ 0 * eta1 + 1 * eta2 + 2 * eta3
    # Variance-covariances of intercepts and slopes
    i ~~ phi1 * i
    s ~~ s
    i ~~ s
    # Means of level (fixed to zero for identification) and slope
    i ~ 0 * 1
    s ~ 1
    # Fixed disturbances of latent outcomes to zero
    eta1 ~ 0 * 1
    eta2 ~ 0 * 1
    eta3 ~ 0 * 1
    # Constrain total variance at Time 0 to 1
    eta1 ~~ psi1 * eta1
    psi1 + phi1 == 1
"
# Use lavaan::sem() to fit a second-order growth model
pstrong_growth_fit <- sem(
  pstrong_growth_mod,
  data = midus_neurotic
)
```

### Table comparing AwC growth model and partial strong invariance 2nd order growth model

The growth parameter estimates of the two models are very similar. 

```{r msummary}
msummary(
  list(
    "AwC growth" = awc_growth_fit2,
    "Partial strong invariance" = pstrong_growth_fit
  ),
  output = "markdown",
  statistic = "conf.int",
  coef_map = c(
    "i ~1 " = "Mean(Level)",
    "s ~1 " = "Mean(Slope)",
    "i ~~ i" = "Var(Level)",
    "s ~~ s" = "Var(Slope)",
    "i ~~ s" = "Cov(Level, Slope)"
  )
)
```

\newpage

## Version Information

```{r, include=FALSE}
options(width = 80)
```

```{r sinfo}
sessionInfo()
```

